---
title: "Estimation of Mixed-Effects Models"
author: "Øystein Sørensen"
format:
  revealjs:
    include-in-header: mathjax-svg.html
bibliography: references.bib
---

```{r, message=FALSE}
library(tidyverse)
```


# Linear Mixed Models


```{r}
#| echo: false
set.seed(11)

# Population parameters: E(y) = b0 + b1 * x
b0 <- 20
b1 <- 0.5
sd_u0 <- 3     # random intercept SD
sd_u1 <- 0.3   # random slope SD
sd_eps <- 1    # residual SD
n_id <- 100
max_followup <- 5

# Irregular measurement times: first visit at 0, last at end of follow-up,
# any visits in between drawn uniformly
sim_times <- function(n_obs, followup) {
  if (n_obs == 1) return(0)
  c(0, sort(runif(n_obs - 2, 0, followup)), followup)
}

dat <- tibble(
  id = factor(seq_len(n_id)),
  n_obs = sample(1:5, n_id, replace = TRUE),
  followup = if_else(n_obs > 1, runif(n_id, 0, max_followup), 0),
  u0 = rnorm(n_id, sd = sd_u0),
  u1 = rnorm(n_id, sd = sd_u1)
) |>
  mutate(
    followup = followup / max(followup) * max_followup, # longest follow-up exactly 8 years
    baseline = runif(n_id, 0, 15 - followup),
    time = map2(n_obs, followup, sim_times)
  ) |>
  unnest(time) |>
  mutate(
    x = baseline + time,
    y = b0 + u0 + (b1 + u1) * x + rnorm(n(), sd = sd_eps)
  )
```

## Data

```{r}
#| echo: false

p <- ggplot(dat, aes(x, y, group = id)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(x = "Time (years)", y = "y") +
  theme_minimal(base_size = 18)

p
```

## Clustered Data

```{r}
#| echo: false

p + geom_line(alpha = 0.4)
```

## Linear Mixed Model

$$
y_{ij} = \beta_{0} + b_{0i} + (\beta + b_{i}) x_{ij} + \epsilon_{ij}
$$

. . .

in matrix form

$$
y = X \beta + Z b + \epsilon
$$

. . .

in `lme4`

```{r}
#| echo: true
library(lme4)
mod <- lmer(y ~ x + (x | id), data = dat)
```

## Fixed Effect Matrix

$$
y = X \beta + Z b + \epsilon
$$

```{r}
#| echo: true

X <- getME(mod, "X")
dim(X)
X[1:8, ]
```


## Random Effect Matrix

$$
y = X \beta + Z b + \epsilon
$$

```{r}
#| echo: true

Z <- getME(mod, "Z")
dim(Z)
Z[1:8, 1:6]
```

## Estimation

Given $b \sim N(0, G)$ and $\epsilon \sim N(0, \sigma^{2} I)$, $\hat{\beta}$ and $\hat{b}$ solve the penalized least squares problem

$$
\min_{\beta,\, b} \; \lVert y - X\beta - Zb \rVert^{2} + \sigma^{2}\, b^{T} G^{-1} b
$$

## Estimation

Henderson's mixed model equations [@hendersonBestLinearUnbiased1975]

$$
\begin{bmatrix}
X^{T}X & X^{T}Z \\
Z^{T}X & Z^{T}Z + \sigma^{2} G^{-1}
\end{bmatrix}
\begin{bmatrix}
\hat{\beta} \\
\hat{b}
\end{bmatrix}
=
\begin{bmatrix}
X^{T}y \\
Z^{T}y
\end{bmatrix}
$$

. . .

- The ridge penalty on $b$ makes the system nonsingular
- But the penalty depends on $G$ and $\sigma^{2}$, which are unknown

## Marginal Likelihood

Integrate out the random effects

$$
L(\beta, \sigma^{2}, G) = \int_{\mathbb{R}^{q}}
\underbrace{p(y \mid b;\, \beta, \sigma^{2})}_{N(X\beta + Zb,\; \sigma^{2} I)}\,
\underbrace{p(b;\, G)}_{N(0,\; G)}
\; \mathrm{d}b
$$

. . .

Closed form: $y \sim N(X\beta,\, ZGZ^{T} + \sigma^{2} I)$

. . .

- $b$ is integrated out: maximize over $\beta$, $\sigma^{2}$ and $G$ alone

## What About Missing Data? {.smaller}

With one grouping factor, the likelihood is a product over participants

$$
L(\beta, \sigma^{2}, G) = \prod_{i=1}^{m} \int p(y_{i} \mid b_{i};\, \beta, \sigma^{2})\, p(b_{i};\, G)\, \mathrm{d}b_{i}
$$

with $y_{i}$ the $n_{i} \geq 1$ observations participant $i$ has

. . .

- We integrate over unobserved $b_{i}$, not over missing $y$: no imputation
- All `r nobs(mod)` observations are used, including `r sum(table(dat$id) == 1)` participants with one measurement
- Valid if missing at random given the model [@rubinInferenceMissingData1976], e.g., dropout depending on earlier observed $y$

## Where Does the Misconception Come From?

- Early software used the EM algorithm, which treats $b$ as missing data and fills in its conditional expectations in the E-step [@lairdRandomEffectsModelsLongitudinal1982]
- Predicted random effects $\hat{b}_{i}$ look like imputed values
- Missing covariates are different: `lmer()` drops those rows, losing their observed $y$, so multiple imputation may be needed

## Maximum Marginal Likelihood

With $\theta = (\sigma^{2}, G)$ and $V_{\theta} = ZGZ^{T} + \sigma^{2} I$

$$
\begin{aligned}
&-2 \ell_{\text{ML}}(\beta, \theta) = \\
& n \log 2\pi + \log \lvert V_{\theta} \rvert + (y - X\beta)^{T} V_{\theta}^{-1} (y - X\beta)
\end{aligned}
$$

. . .

For given $\theta$, the maximizer over $\beta$ is the GLS estimate

$$
\hat{\beta}_{\theta} = \left(X^{T} V_{\theta}^{-1} X\right)^{-1} X^{T} V_{\theta}^{-1} y
$$

. . .

- Plug in $\hat{\beta}_{\theta}$ and optimize the profile likelihood over $\theta$ only

## Restricted Maximum Likelihood

Integrate out $\beta$ too [@harvilleBayesianInferenceVariance1974]

$$
L_{\text{REML}}(\theta) = \int_{\mathbb{R}^{p}} L(\beta, \theta) \, \mathrm{d}\beta
$$

. . .

$$
\begin{aligned}
& -2 \ell_{\text{REML}}(\theta) = \\
&-2 \ell_{\text{ML}}(\hat{\beta}_{\theta}, \theta) + \log \lvert X^{T} V_{\theta}^{-1} X \rvert - p \log 2\pi
\end{aligned}
$$

. . .

- ML ignores that $\beta$ is estimated: variance components biased downward, like $\text{RSS}/n$ vs. $\text{RSS}/(n - p)$
- Matters with few clusters or many fixed effects

## REML and Model Comparison

REML likelihoods are not comparable

- Likelihood ratio tests, AIC, BIC involving fixed effects: use ML
  - `anova()` in `lme4` refits REML models with ML automatically
- Same fixed effects, different random effects: REML is fine

## Inference in Practice {.smaller}

Fixed effects

- `lmer()` reports no p-values: with unbalanced data the null distribution of the $t$ statistic, and its degrees of freedom, are not known exactly
- Small samples: Satterthwaite (`lmerTest`) or Kenward–Roger (`pbkrtest`) degrees of freedom, based on REML [@kenwardSmallSampleInference1997]

Confidence intervals

- `confint(mod)` gives profile likelihood intervals, also for the standard deviations and correlations; Wald intervals are poor for variance parameters
- Parametric bootstrap when in doubt: `confint(mod, method = "boot")`.

## Testing Variance Components {.smaller}

- $H_{0}: \sigma_{b}^{2} = 0$ is on the boundary, so the LRT statistic is $\tfrac{1}{2}\chi^{2}_{0} + \tfrac{1}{2}\chi^{2}_{1}$, not $\chi^{2}_{1}$: halve the naive p-value
- Adding a correlated random slope: $\tfrac{1}{2}\chi^{2}_{1} + \tfrac{1}{2}\chi^{2}_{2}$
- The mixtures are asymptotic; with few clusters, use the parametric bootstrap

. . .

Is there a random slope for $x$? Same fixed effects, so REML is fine

```{r}
#| echo: true
mod0 <- lmer(y ~ x + (1 | id), data = dat)
lrt <- as.numeric(2 * (logLik(mod) - logLik(mod0)))
c(naive = pchisq(lrt, df = 2, lower.tail = FALSE),
  mixture = mean(pchisq(lrt, df = 1:2, lower.tail = FALSE)))
```

# Generalized Linear Mixed Models

## Marginal Likelihood

$y$ distributed according to an exponential family with mean $\mu$, and linear predictor

$$
g(\mu) = \eta = X \beta + Z b
$$

. . .

Integral no longer known

$$
L(\beta, \phi, G) = \int_{\mathbb{R}^{q}}
\underbrace{p(y \mid b;\, \beta, \phi)}_{\text{exp. fam.}}\,
\underbrace{p(b;\, G)}_{N(0,\; G)}
\; \mathrm{d}b
$$

## Common Integral Approximations

:::{.incremental}
- Penalized quasi-likelihood [@breslowApproximateInferenceGeneralized1993]: very fast but crude. Implemented in `MASS::glmmPQL()`.
- Laplace approximation [@tierneyAccurateApproximationsPosterior1986]: very good for $\beta$ but typically biased for $G$. Implemented in `lme4::glmer()` and `glmmTMB::glmmTMB()`.
- Gaussian quadrature [@pinheiroEfficientLaplacianAdaptive2006]: useful if $G$ is a scalar and you care deeply about it. 
:::

## Laplace Approximation

Write the integrand as $e^{h(b)}$, with

$$
h(b) = \log p(y \mid b;\, \beta, \phi) + \log p(b;\, G)
$$

. . .

Second-order Taylor expansion of the exponent around the mode $\hat{b}$, where $\nabla h(\hat{b}) = 0$:

$$
h(b) \approx h(\hat{b}) - \frac{1}{2} (b - \hat{b})^{T} H (b - \hat{b})
$$

with $H = -\nabla^{2} h(\hat{b}) = Z^{T} W Z + G^{-1}$ and $W$ the diagonal GLM weights at $\hat{b}$

## Laplace Approximation

The integrand is now a Gaussian kernel

$$
\begin{aligned}
L(\beta, \phi, G) &\approx e^{h(\hat{b})} \int_{\mathbb{R}^{q}} e^{-\frac{1}{2} (b - \hat{b})^{T} H (b - \hat{b})} \, \mathrm{d}b \\
&= e^{h(\hat{b})} \, (2\pi)^{q/2} \, \lvert H \rvert^{-1/2}
\end{aligned}
$$

. . .

Taking $-2 \log$:

$$
\begin{aligned}
-2 \ell(\beta, \phi, G) \approx\; & \underbrace{-2 \log p(y \mid \hat{b};\, \beta, \phi) + \hat{b}^{T} G^{-1} \hat{b}}_{\text{penalized deviance}} \\
& + \log \lvert G \rvert + \log \lvert Z^{T} W Z + G^{-1} \rvert
\end{aligned}
$$

## Same Form as the LMM

For the LMM, $h(b)$ is exactly quadratic in $b$, so the formula is exact, with

$$
-2 \log p(y \mid \hat{b};\, \beta, \sigma^{2}) = n \log 2\pi\sigma^{2} + \frac{\lVert y - X\beta - Z\hat{b} \rVert^{2} }{\sigma^{2}}
$$

$W = I / \sigma^{2}$, and $\hat{b}$ the penalized least squares solution from before

. . .

- GLMM: approximate, with $\hat{b}$ the conditional mode
- Accurate when $p(b \mid y)$ is close to Gaussian; worse with few, binary observations per cluster

## Finding the Conditional Mode {.smaller}

Penalized iteratively reweighted least squares (PIRLS) for given $\beta$, $G$, from $b^{(0)} = 0$:

1. $\eta = X\beta + Zb^{(k)}$, $\mu = g^{-1}(\eta)$
2. $z_{i} = (Zb^{(k)})_{i} + (y_{i} - \mu_{i})\, g'(\mu_{i})$, $\; w_{i} = 1 / \{\phi\, V(\mu_{i})\, g'(\mu_{i})^{2}\}$
3. Solve $\left(Z^{T} W Z + G^{-1}\right) b^{(k+1)} = Z^{T} W z$
4. Repeat until the penalized deviance converges

. . .

- Each step is the LMM problem, with weights $W$ and response $z$
- Runs inside every evaluation of the Laplace likelihood; an outer optimizer handles $\beta$ and $G$

## Conditional vs. Marginal Effects {.smaller}

$$
E(y \mid b) = g^{-1}(X\beta + Zb), \qquad
E(y) = \int g^{-1}(X\beta + Zb)\, p(b)\, \mathrm{d}b \neq g^{-1}(X\beta)
$$

- Identity link: the two coincide, so $\beta$ is also the population-averaged effect
- Other links: $\beta$ is a subject-specific effect, and the population-averaged curve is flatter

```{r}
#| echo: false
#| fig-height: 4
# Logistic random intercept model with slope 1 and random intercept SD 2
sd_b_marg <- 2
x_grid <- seq(-6, 6, length.out = 200)

patients <- expand_grid(b = qnorm(ppoints(15)) * sd_b_marg, x = x_grid) |>
  mutate(p = plogis(x + b), curve = "Individual patients")

marginal_p <- function(eta) {
  integrate(\(b) plogis(eta + b) * dnorm(b, sd = sd_b_marg), -Inf, Inf)$value
}

curves <- tibble(x = x_grid) |>
  mutate(
    `Typical patient (b = 0)` = plogis(x),
    `Population average` = map_dbl(x, marginal_p)
  ) |>
  pivot_longer(-x, names_to = "curve", values_to = "p")

ggplot(mapping = aes(x, p, colour = curve)) +
  geom_line(data = patients, aes(group = b), linewidth = 0.5) +
  geom_line(data = curves, linewidth = 1.5) +
  scale_colour_manual(
    values = c(`Individual patients` = "grey75", `Typical patient (b = 0)` = "#2a78d6",
               `Population average` = "#eb6834"),
    breaks = c("Individual patients", "Typical patient (b = 0)", "Population average")
  ) +
  labs(x = "x", y = "P(y = 1)", colour = NULL) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Probit vs. Logit {.smaller}

Random intercept $b \sim N(0, \sigma_{b}^{2})$, linear predictor $\eta = x^{T}\beta$, and $\varepsilon \sim N(0, 1)$ independent of $b$

Population-averaged effect with Probit:

$$
E\{\Phi(\eta + b)\} = P(\varepsilon - b \leq \eta) = \Phi\left(\eta \big/ \sqrt{1 + \sigma_{b}^{2}}\right)
$$

Logit, approximate, using $\operatorname{logit}^{-1}(t) \approx \Phi(t / 1.7)$:

$$
E\{\operatorname{logit}^{-1}(\eta + b)\} \approx \operatorname{logit}^{-1}\left(\eta \big/ \sqrt{1 + 0.346\, \sigma_{b}^{2}}\right)
$$

. . .

- Probit: the marginal model is again a probit model, with coefficients $\beta / \sqrt{1 + \sigma_{b}^{2}}$
- Logit: the GLMM odds ratio is a within-patient odds ratio, not a population odds ratio
- With random slopes the scaling depends on the covariates, for both links
- Population-averaged effects: average predictions over $b$, or fit GEE

# Numerics and Implementation Details


## Sparse Matrices

Imagine we follow students over time, nested within schools, and that students may change school.

```{r}
#| echo: false
# Add school membership to dat: students start in a random school and
# move to a new random school at each later visit with probability 0.3
set.seed(12)
n_school <- 40
school_eff <- rnorm(n_school, sd = 2)

dat_school <- dat |>
  group_by(id) |>
  mutate(school = accumulate(
    runif(n() - 1) < 0.3,
    \(s, move) if (move) sample(n_school, 1) else s,
    .init = sample(n_school, 1)
  )) |>
  ungroup() |>
  mutate(y = y + school_eff[school], school = factor(school))
```

```{r}
#| echo: true
mod_school <- lmer(y ~ x + (x | id) + (1 | school), data = dat_school)
```

. . .

In `nlme`, crossed effects need a dummy grouping factor spanning all observations

```{r}
#| echo: true
#| eval: false
dat_school$all <- factor(1)
lme(y ~ x, data = dat_school,
    random = list(all = pdIdent(~ 0 + school), id = ~ x))
```

## Sparse Matrices in lme4

```{r}
#| echo: false
#| fig-height: 5
# lme4: only the nonzero entries of Z are stored
Z <- getME(mod_school, "Z")
n_id_cols <- 2 * nlevels(dat_school$id)

lme4_df <- as.data.frame(summary(Z)) |>
  as_tibble() |>
  transmute(row = i, col = j, value = if_else(j <= n_id_cols, "Student", "School"))

# nlme: student intercept and slope columns, plus one dense indicator
# column per school, with every zero stored
nlme_df <- expand_grid(
  row = seq_len(nrow(dat_school)),
  col = seq_len(2 + nlevels(dat_school$school))
) |>
  mutate(value = case_when(
    col <= 2 ~ "Student",
    as.integer(dat_school$school)[row] == col - 2 ~ "School",
    .default = "Stored zero"
  ))

labs_pkg <- c(
  lme4 = str_glue("lme4, sparse: {scales::comma(nrow(lme4_df))} values stored"),
  nlme = str_glue("nlme, dense: {scales::comma(nrow(nlme_df))} values stored")
)

bind_rows(lme4 = lme4_df, nlme = nlme_df, .id = "package") |>
  mutate(package = labs_pkg[package]) |>
  ggplot(aes(col, row, fill = value)) +
  geom_raster() +
  facet_grid(~ package, scales = "free_x") +
  scale_y_reverse(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_fill_manual(
    values = c(Student = "#2a78d6", School = "#eb6834", `Stored zero` = "grey88"),
    breaks = c("Student", "School", "Stored zero")
  ) +
  labs(x = "Column", y = "Observation", fill = NULL) +
  theme_minimal(base_size = 18) +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom",
    panel.border = element_rect(fill = NA, colour = "grey60")
  )
```


## Packages

- `lme4` [@batesFittingLinearMixedEffects2015]: sparse matrices; Laplace approximation; probably the fastest?
- `glmmTMB` [@brooksGlmmTMBBalancesSpeed2017]: sparse matrices; Laplace approximation; automatic differentiation; easily handles lots of different distributions.
- `brms` [@burknerBrmsPackageBayesian2017]: `lme4` syntax; Stan backend; generates Stan code; very flexible.
- `nlme`: historically important; maintained but not actively developed.



# Numerical Instability

## Singular Fits {.smaller}

`boundary (singular) fit: see help('isSingular')`

- The estimate is on the boundary: a variance of zero or a correlation of $\pm 1$
- A property of the likelihood for these data, not a software bug
- Typical with few clusters, few observations per cluster, or many variance parameters

. . .

Henderson's equations need $G^{-1}$, which then does not exist. `lme4` instead writes $b = \Lambda_{\theta} u$ with $u \sim N(0, \sigma^{2} I)$ [@batesFittingLinearMixedEffects2015]

$$
\min_{\beta,\, u} \; \lVert y - X\beta - Z\Lambda_{\theta} u \rVert^{2} + \lVert u \rVert^{2}
$$

so a singular $\Lambda_{\theta}$ causes no problems

## Convergence Warnings {.smaller}

`Model failed to converge with max|grad| = 0.006 (tol = 0.002)`

- The outer optimizer over $\theta$ (and $\beta$ for GLMMs) stopped where the gradient is not quite zero
- Often a false positive, especially in large data sets (see `?lme4::convergence`)
- Check: `allFit(mod)` refits with all available optimizers; if the estimates agree, report them

. . .

`Model is nearly unidentifiable: large eigenvalue ratio - Rescale variables?`

- Center and scale covariates, e.g., time in decades or centered at its mean
- Don't ask for more random effects than the data can support

## Binary Outcomes in Small Data {.smaller}

40 patients with 2 binary measurements each, true random intercept SD 1.5, 200 simulated data sets

```{r}
#| echo: false
#| cache: true
# Estimated random intercept SD, or NA if the fit fails
glmer_sd <- possibly(function(d, ...) {
  f <- suppressWarnings(suppressMessages(
    glmer(y ~ time + (1 | id), data = d, family = binomial, ...)
  ))
  sqrt(VarCorr(f)$id[1, 1])
}, otherwise = NA_real_)

set.seed(2026)
n_pat <- 40
sd_b_true <- 1.5

sim_bin <- map(1:200, \(rep) {
  d <- tibble(id = factor(rep(seq_len(n_pat), each = 2)), time = rep(0:1, n_pat)) |>
    mutate(y = rbinom(n(), 1, plogis(-0.5 + 0.5 * time + rnorm(n_pat, sd = sd_b_true)[id])))
  tibble(
    method = c("Laplace", "Adaptive quadrature"),
    sd_b = c(glmer_sd(d), glmer_sd(d, nAGQ = 25))
  )
}) |>
  list_rbind() |>
  mutate(method = fct_inorder(method))

sim_summary <- sim_bin |>
  group_by(method) |>
  summarise(
    median_sd = median(sd_b, na.rm = TRUE),
    pct_zero = mean(sd_b < 1e-4, na.rm = TRUE)
  )
```

```{r}
#| echo: false
#| fig-height: 4.2
sim_bin |>
  filter(!is.na(sd_b)) |>
  mutate(type = if_else(sd_b < 1e-4, "Singular fit", "Positive estimate")) |>
  ggplot(aes(sd_b, fct_rev(method))) +
  geom_jitter(aes(colour = type), height = 0.25, width = 0, alpha = 0.7, size = 2.5) +
  geom_point(data = sim_summary, aes(median_sd), shape = "|", size = 16) +
  geom_vline(xintercept = sd_b_true, linetype = "dashed") +
  annotate("text", x = sd_b_true, y = 2.45, label = "true value", hjust = -0.1, size = 5) +
  scale_colour_manual(values = c(`Positive estimate` = "#2a78d6", `Singular fit` = "#eb6834")) +
  labs(x = "Estimated random intercept SD (vertical bar: median)", y = NULL, colour = NULL) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

Laplace underestimates the SD (median `r round(sim_summary$median_sd[1], 2)` vs. `r round(sim_summary$median_sd[2], 2)` with quadrature). Singular fits occur with both methods (Laplace `r scales::percent(sim_summary$pct_zero[1])`, quadrature `r scales::percent(sim_summary$pct_zero[2])`): they come from the likelihood, not the approximation.

## What Can We Do?

- Treat a singular fit as information: these data cannot distinguish the variance from zero
- Simplify the random effects: drop correlations with `(x || id)`, or drop random slopes
- More accurate integral: `glmer(..., nAGQ = 25)`, available with a single random intercept
- Bayesian fit with `brms`: the posterior shows the full uncertainty about the SD, instead of a point estimate at zero


## Take-Home Messages

- No imputation: the random effects are integrated out
- Missing outcomes: fine under MAR; missing covariates: not
- REML for variances; ML to compare fixed effects
- Small binary data: check Laplace against quadrature
- GLMM coefficients are subject-specific
- Singular fits and warnings tell you about the data

# Questions? Comments?


## References